---
output:
  html_document:
    keep_md: yes
  pdf_document: default
---
### Exercise quality prediction  

#### Dan Wong  
  
  
#### Overview  

This document details the creation, training and selection of a machine learning algorithm to classify the quality of exercise (weight lifting) using accelerometer data.

Produced as part of the Coursera Practical Machine learning course, week 4 assignment.  

Data, methodology and the original paper can be found at the link below, courtesy of;  
Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H.  
[Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements](http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/work.jsf?p1=10335)


#### Data load, set up and initialisation
```{r echo=FALSE}

###### THESE commands should be removed from the document once complete
setwd("~/Study/Machine_Learning/Week_4_Project")
training <- read.csv("pml-training.csv", na.strings=c("NA","#DIV/0!",""))
testing <- read.csv("pml-testing.csv", na.strings=c("NA","#DIV/0!",""))


###### REMOVE THE ABOVE COMMANDS ON COMPLETION

# initialise variables and set seed to allow reproduction
set.seed(696969)

#load the data

########## UNCOMMENT THESE COMMANDS ON COMPLETION
## trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
## testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
## training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
## testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
########## UNCOMMENT THESE COMMANDS ON COMPLETION

#references
library(caret)
library(rpart)
library(randomForest)

```


#### Preprocessing data  

Examination of the data shows various redundant columns that seem unlikely to be useful for classification purposes (indices and timestamps).  
Further many columns contain a large number (>95%) of NA values.  
Whilst it is possible that some of these variables have some prediction utility - they have been removed here in the interests of brevity and simplicity.

```{r echo=FALSE}

# check the number of na's in each column 
redundant <- data.frame(colSums(is.na(training)))
redundant <- cbind(redundant,rownames(redundant))
colnames(redundant)<- c("na_count", "column_name")
redundant <- redundant[order(-redundant$na_count),]

# select the column names where 95% of the entries are na
remove <- subset(redundant, na_count >0.95*nrow(training))
remove <- as.vector(remove[,2])

# add some other manually selected column names to remove 
# (indices, timestamps and usernames)
remove[101:106] <- c("X","raw_timestamp_part_1","raw_timestamp_part_2","cvtd_timestamp","user_name","new_window")

#remove these columns from the training and test data
training <- training[ , !(names(training) %in% remove)]
testing <- testing[ , !(names(testing) %in% remove)]
```


#### Model creation and training
  
Training data is partitioned into a training and validation set to allow the estimation of out of sample error.

```{r echo=TRUE}
# create partition and split data
inTrain <- createDataPartition(y=training$classe,
                              p=0.7, list=FALSE)
training_data <- training[inTrain,]
validation_data <- training[-inTrain,]
```

A random forest model is trained using 5 fold cross validation.

```{r echo=TRUE}
# specify the cross validation method
fitControl <- trainControl(method = "cv",
                           number = 5)
# Train the model
#rf_model_non_pca <- train(classe ~.,data=training_data,method="rf",trControl=fitControl,prox=TRUE)

rf_model_non_pca

# graph the variables
varImpPlot(rf_model_non_pca$finalModel,type=2)

```

An out of sample accuracy estimate is provided by applying this model to the validation data set.

```{r echo=TRUE}
# predict values and create confusion matrix
confusionMatrix(predict(rf_model_non_pca, newdata = validation_data), validation_data$classe)

#tidy up the workspace
rm(training_data)
rm(validation_data)
rm(rf_model_non_pca)
gc()

```

Since accuracy looks acceptable a similar model will be trained using the entire training dataset and used to predict on the assigned testing data.

```{r echo=TRUE}
# Train the final model
final_model <- train(classe ~.,data=training,method="rf",trControl=fitControl,prox=TRUE)

# Predict the testing values
predict(final_model,newdata=testing)

```